Mood increases during video game play: A naturalistic observational case study of PowerWash Simulator

### 2023-09-06 ms.qmd
WIP changes

library(scales)
library(cmdstanr)
library(posterior)
library(brms)
library(ggstance)
library(patchwork)
library(tidyverse)
```

@@

## Measures

We measured affective well-being with one item.

@@

More formally, we modelled each mood response *i* in 1 to `r s$n_obs` from participant *j* in 1 to `r s$n_sub` and participant-session *k* in 1 to `r s$n_ses` with a normal distribution censored at the VAS limits [0, 1]. We then modelled the distribution's mean ($\eta$) on a population-level intercept ($\beta_0$) and contrast of time ($\beta_1$), their random per-player deviations $\alpha$, and a random intercept for player-sessions $\gamma_0$:

@@

We analyzed the data with R [@rcoreteamLanguageEnvironmentStatistical2023] and used the brms package to estimate, via Stan's HMC sampling procedures, and post-process the models [@@burknerBrmsPackageBayesian2017; @standevelopmentteamStanModelingLanguage2022]. We drew `r number2(ndraws(fit))` samples from the posterior distribution using brms' default prior distributions on all parameters, and used numerical and graphical checks to ensure model convergence and adequacy.

@@

## Descriptives

```{r}
#| label: fig-descriptives
#| include: true
#| fig-height: 2.2
#| fig-width: 7
#| fig-cap: '**A**. Histogram of session durations (note log10 x-axis). **B**. Summary of how many sessions each participant completed. **C**. Histogram of all mood ratings.'

@@

p_durations <- p0 %+%
  dat_session_durations +
  aes(duration) +
  scale_x_log10() +
  labs(x = "Session duration (minutes)", y = "Sessions")

@@

```{r}
probs <- c(.5, .1, .9)
tmp <- list()
tmp$a <- quantile(dat_session_durations$duration, probs = probs, na.rm = TRUE)
tmp$b <- quantile(dat_session_counts$sessions, probs = probs, na.rm = TRUE)
tmp$c <- quantile(dat$mood, probs = probs, na.rm = TRUE)
tmp$d <- quantile(dat$mood[dat$post == 0], probs = probs, na.rm = TRUE)
tmp$e <- quantile(dat$mood[dat$post == 1], probs = probs, na.rm = TRUE)
```

The median session duration was `r tmp$a` minutes [10 and 90 percentiles] (@fig-descriptives A); the median player contributed data from `r tmp$b` sessions (@fig-descriptives B), and the median mood was `r tmp$c` (@fig-descriptives C; pre-session: `r tmp$d`, during play: `r tmp$e`).

```{r}
#| label: fig-data
#| include: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "**A**. Representative data from three participants (rows) showing scatterplots of mood responses over that session's (columns) duration. Each session's first mood report is highlighted in red, subsequent reports are in blue. **B**. Horizontal histograms of all sessions' mood response for that person. **C**. Scatterplot of person-mean mood reports at the beginning of the session (x-axis) and subsequent reports (y-axis). **D**. Histograms of mood ratings at the beginning of session, during the session, and differences between person-means of first and subsequent mood reports."

@@

# Take a quasirandom sample of people who had sufficient number of sessions
# and data per session.
set.seed(99)
tmp <- dat |>
  add_count(pid, session, name = "obs_per_session") |>
  filter(obs_per_session >= 4) |>
  mutate(session = as.numeric(as.factor(session)), .by = pid) |>
  mutate(session_per_person = length(unique(session)), .by = pid) |>
  filter(session_per_person >= 4) |>
  arrange(pid, session) |>
  filter(pid %in% sample(unique(pid), 3)) |>
  mutate(Session = str_glue("Session {session}"), Person = str_glue("Person {fct_anon(pid)}"))

library(ggdist)
p_mood_duration_sessions <- tmp |> 
  ggplot() +
  aes(duration / 60, mood, col = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_continuous(
    "Session duration (hours)"
  ) +
  scale_y_continuous(
    "Mood",
    limits = c(0, 1),
    breaks = axislabels,
    labels = axislabels
  ) +
  geom_point(size = 1.5, alpha = 1) +
  facet_grid(
    rows = vars(Person), 
    cols = vars(Session),
    scales = "free_x"
  ) +
  theme(
    legend.position = "none",
    strip.text.y = element_blank()
  )

p_mood_duration_all <- tmp |> 
  mutate(Session = "All sessions") |> 
  ggplot() +
  aes(y = mood, col = post, fill = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_continuous(
    "Scaled count",
    expand = expansion(c(0.01, 0.1))
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = axislabels,
    labels = axislabels
  ) +
  geom_dotplot(
    aes(x = 0),
    dotsize = 0.9,
    binaxis = "y",
    method = "histodot",
    binwidth = 0.06,
    stackgroups = TRUE
  ) +
  facet_grid(
    rows = vars(Person), 
    cols = vars(Session)
  ) +
  theme(
    axis.text = element_blank(), 
    axis.title.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    strip.background.y = element_blank(),
    legend.position = "none"
  )

p_mood_duration <- (p_mood_duration_sessions | p_mood_duration_all) +
  plot_layout(widths = c(5/6, 1/6))

tmp <- dat |> 
  summarise(
    mood = mean(mood, na.rm = TRUE),
    .by = c(pid, post)
  ) |> 
  pivot_wider(names_from = post, values_from = mood) |> 
  mutate(delta = `1` - `0`)

p_mood_biscatter <- tmp |> 
  ggplot(aes(`0`, `1`)) +
  scale_x_continuous(
    "Pre-session mood",
    expand = expansion(0.01),
    breaks = axislabels,
    labels = axislabels
  ) +
  scale_y_continuous(
    "Mood during session",
    expand = expansion(0.01),
    breaks = axislabels,
    labels = axislabels
  ) +
  geom_point(
    alpha = .25, size = 1, shape = 1
  ) +
  geom_abline(linewidth = .33, col = "dodgerblue") +
  theme(aspect.ratio = 1)

p_mood_difference <- tmp |> 
  rename(
    `Pre-session` = `0`,
    `During session` = `1`,
    `Difference (During - Pre)` = delta
  ) |> 
  pivot_longer(-pid) |> 
  mutate(name = fct_inorder(name)) |> 
  ggplot(aes(value)) +
  geom_histogram(bins = 30, col = "white") +
  scale_y_continuous(
    "Participants",
    expand = expansion(c(0.001, 0.1))
  ) +
  scale_x_continuous(
    "Mood",
    expand = expansion(c(0.001))
  ) +
  facet_wrap("name", ncol = 3, scales = "free")

(p_mood_duration / (p_mood_biscatter | p_mood_difference)) +
  plot_layout(heights = c(6, 4)) +
  plot_annotation(tag_level = "A")
```

@@

We then turned to our primary research question: To what extent does player mood in PWS change from pre-play to during play? First, we visualize the relevant data in (@fig-data): Panel A shows mood responses from three example participants' first five sessions of play. @fig-data B then shows histograms of those data to facilitate visual comparison of the raw data. We then calculated participants' mean moods at the beginning and during each session (@fig-data C): Visual inspection shows a considerable shift toward higher values for the during play means. @fig-data D shows histograms of the pre- and during play mood means, and their difference, and suggests that while many differences are small, overall mean moods appeared greater during play than pre-play.

@@

```{r}
#| label: scaled-coefficients
#| eval: false

library(latex2exp)
tmp_ <- as_draws_df(
  fit, 
  c("sd_pid__Intercept", "sd_pid__post1", "sigma", "b_post1")
) |> 
  mutate(
    total_residual = sd_pid__Intercept + sd_pid__post1 + sigma,
    z0 = b_post1 / total_residual,
    z1 = b_post1 / sd_pid__Intercept,
    z2 = b_post1 / sd_pid__post1,
    z3 = b_post1 / sigma
  )

tmp <- tmp_ |> 
  summarise_draws(
    mean = ~number(mean(.), .001), 
    ~number2(quantile2(., probs = c(.025, .975)), .001)
  ) |> 
  mutate(
    x = str_glue("{mean} [{q2.5}, {q97.5}]")
  )
tmp_ |> 
  select(b_post1, starts_with("z")) |> 
  pivot_longer(everything()) |> 
  ggplot(aes(value, name)) +
  stat_slabinterval(
    .width = c(.8, .95),
    normalize = "xy",
    height = .5,
    interval_size_range = c(0.5, 1),
    fatten_point = 1
  ) +
  scale_x_continuous(
    "Contrast magnitude",
    breaks = scales::extended_breaks(7)
  ) +
  scale_y_discrete(
    labels = c(
      TeX(r"($\alpha_1$)"),
      TeX(r"($\frac{\alpha_1}{\tau_0 + \tau_1 + \sigma}$)"),
      TeX(r"($\frac{\alpha_1}{\tau_0}$)"),
      TeX(r"($\frac{\alpha_1}{\tau_1}$)"),
      TeX(r"($\frac{\alpha_1}{\sigma}$)")
    ),
    expand = expansion(0.1)
  ) +
  theme(
    axis.title.y = element_blank()
  )

coef(fit, summary = FALSE)$pid[,,"post1"] |> 
  as_tibble()

fixef(fit, summary = FALSE)[,"post1", drop = FALSE] |> 
  as_tibble()
```

To further interpret the magnitude of the difference in moods between pre- and during play, we compared the contrast to the natural variations in mood as estimated by the model. Specifically, we compared the estimated pre - during contrast to the standard deviation of the between-person differences in the intercept ($\tau_{\alpha_0}$ = `r tmp$x[1]`), the contrast ($\tau_{\alpha_1}$ = `r tmp$x[2]`) and to the residual within-person variation ($\sigma$).

>>>>

library(scales)
library(cmdstanr)
library(posterior)
library(brms)
library(ggstance)
library(ggdist)
library(patchwork)
library(tidyverse)
```

@@

@@

More formally, we modelled each mood response *i* in 1 to `r s$n_obs` from participant *j* in 1 to `r s$n_sub` and participant-session *k* in 1 to `r s$n_ses` with a normal distribution censored at the VAS limits [0, 1]. We then modelled the distribution's mean ($\eta$) on a population-level intercept ($\beta_0$) and contrast-coded measurement timepoint ($\beta_1$), their random per-player deviations $\alpha$, and a random intercept for player-sessions $\gamma_0$:

@@

We analyzed the data with R [@rcoreteamLanguageEnvironmentStatistical2023] and used the brms package to estimate, via Stan's HMC sampling procedures, and post-process the models [@burknerBrmsPackageBayesian2017; @standevelopmentteamStanModelingLanguage2022]. We drew `r number2(ndraws(fit))` samples from the posterior distribution using brms' default prior distributions on all parameters, and used numerical and graphical checks to ensure model convergence and adequacy.

@@

```{r}
#| label: fig-descriptives
#| include: false
#| fig-height: 2.2
#| fig-width: 7
#| fig-cap: '**A**. Histogram of session durations (note log10 x-axis). **B**. Summary of how many sessions each participant completed. **C**. Histogram of all mood ratings.'

@@

p_durations <- p0 %+%
  dat_session_durations +
  aes(duration) +
  scale_x_log10() +
  coord_cartesian(xlim = c(5, 900)) +
  labs(x = "Session duration (minutes)", y = "Sessions")

@@

```{r}
tmp <- list(
  dat_session_durations$duration,
  dat_session_counts$sessions,
  dat$mood,
  dat$mood[dat$post == 0],
  dat$mood[dat$post == 1]
) |> 
  map(~median_qi(.x, .width = 0.8, na.rm = TRUE)) |> 
  map(~str_glue_data(.x, "{y} [{ymin}, {ymax}]"))
```

The median session duration was `r tmp[[1]]` minutes [10 and 90 percentiles]; the median player contributed data from `r tmp[[2]]` sessions, and the median mood was `r tmp[[3]]` (pre-session: `r tmp[[4]]`, during play: `r tmp[[5]]`).

```{r}
#| label: fig-data
#| include: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "**A**. Representative data from three participants (rows) showing scatterplots of mood responses over that session's (columns) duration. Each session's first mood response is highlighted in red, subsequent reports are blue. **B**. Horizontal histograms of all sessions' mood response for that person, with means and bootstrapped 95%CIs. **C**. Histograms of session-mean (**D**. person-mean) moods before (top) and during (bottom) sessions. **E**. Differences in session-mean (**F**. player-mean) mood differences (during session - pre-play). **G**. Scatterplot of person-mean mood reports at beginning (x-axis) and during game play sessions (y-axis)."

@@

# Plot: Example raw sessions
set.seed(99)
dat_example_sessions <- dat |>
  add_count(pid, session, name = "obs_per_session") |>
  filter(obs_per_session >= 4) |>
  mutate(session = as.numeric(as.factor(session)), .by = pid) |>
  mutate(session_per_person = length(unique(session)), .by = pid) |>
  filter(session_per_person >= 4) |>
  arrange(pid, session) |>
  filter(pid %in% sample(unique(pid), 3)) |>
  mutate(Session = str_glue("Session {session}"), Person = str_glue("Person {fct_anon(pid)}"))

p_mood_example_sessions <- dat_example_sessions |> 
  ggplot() +
  aes(duration / 60, mood, col = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_continuous(
    "Session duration (hours)"
  ) +
  scale_y_continuous(
    "Mood",
    limits = c(0, 1),
    breaks = axislabels,
    labels = axislabels
  ) +
  geom_point(size = 1.5, alpha = 1) +
  facet_grid(
    rows = vars(Person), 
    cols = vars(Session),
    scales = "free_x"
  ) +
  theme(
    legend.position = "none",
    strip.text.y = element_blank()
  )

p_mood_example_all <- dat_example_sessions |> 
  mutate(Session = "All sessions") |> 
  ggplot() +
  aes(y = mood, x = post, col = post, fill = post) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  scale_x_discrete(
    "Session",
    breaks = c(0, 1),
    labels = c("Pre", "During"),
    expand = expansion(c(0.2, 2.4))
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = axislabels,
    labels = axislabels
  ) +
  stat_summary(
    fun.data = mean_cl_boot,
    fatten = 2
  ) +
  geom_dots(side = "right", alpha = .33, binwidth = .075) +
  facet_grid(
    rows = vars(Person), 
    cols = vars(Session)
  ) +
  theme(
    axis.text.y = element_blank(), 
    axis.title.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    axis.text.x = element_text(hjust = 0),
    strip.background.y = element_blank(),
    legend.position = "none"
  )

p_mood_example <- (p_mood_example_sessions | p_mood_example_all) +
  plot_layout(widths = c(92, 8))

# Plot: Raw mood histograms at pre and during play
p_mood_prepost_raw <- dat |> 
  mutate(post = factor(post, labels =  c("Pre-play", "During play"))) |>
  ggplot(aes(mood, fill = post)) +
  scale_color_brewer(
    "Pre-session measure",
    palette = "Set1",
    aesthetics = c("color", "fill")
  ) +
  geom_histogram(bins = 30, col = "white") +
  scale_y_continuous(
    "Observations",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  ) +
  scale_x_continuous(
    "Mood",
    expand = expansion(c(0.001)),
    breaks = axislabels
  ) +
  facet_wrap("post", ncol = 1, scales = "free_y") +
  theme(legend.position = "none")

# Plot: Person-session-mean mood histograms at pre and during play
p_mood_prepost_sessions <- p_mood_prepost_raw %+%
  summarise(
    p_mood_prepost_raw$data,
    mood = mean(mood, na.rm = TRUE), .by = c(post, pid, ps)
  ) +
  scale_y_continuous(
    "Sessions",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  )

# Plot: Person-mean mood histograms at pre and during play
p_mood_prepost_players <- p_mood_prepost_sessions %+%
  summarise(
    p_mood_prepost_raw$data,
    mood = mean(mood, na.rm = TRUE), .by = c(post, pid)
  ) +
  scale_y_continuous(
    "Players",
    expand = expansion(c(0.001, 0.1)),
    breaks = scales::extended_breaks(5)
  )

# Plot: Difference histogram (sessions)
p_mood_difference_sessions <- p_mood_prepost_sessions$data |> 
  pivot_wider(names_from = post, values_from = mood) |> 
  ggplot(aes(`During play` - `Pre-play`)) +
  geom_histogram(bins = 50, col = "white") +
  geom_vline(xintercept = 0, linewidth = .33, col = "dodgerblue") +
  scale_y_continuous(
    "Sessions",
    expand = expansion(c(0.001, 0.1))
  ) +
  scale_x_continuous(
    "Difference",
    expand = expansion(c(0.001))
  ) +
  coord_cartesian(xlim = c(-0.4, 0.6))

# Plot: Difference histogram (players)
p_mood_difference_players <- p_mood_difference_sessions %+% 
  summarise(
    p_mood_difference_sessions$data,
    across(`Pre-play`:`During play`, ~mean(.x, na.rm = TRUE)),
    .by = pid
  ) +
  scale_y_continuous(
    "Players",
    expand = expansion(c(0.001, 0.1))
  )

p_mood_difference <- (
  (p_mood_difference_sessions + 
     theme(
      axis.title.x = element_blank(), 
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    )) / 
    p_mood_difference_players
  )

p_mood_biscatter_sessions <- p_mood_difference_sessions$data |> 
  ggplot(aes(`Pre-play`, `During play`)) +
  scale_x_continuous(
    expand = expansion(0.01),
    breaks = axislabels,
    labels = axislabels
  ) +
  scale_y_continuous(
    expand = expansion(0.01),
    breaks = axislabels,
    labels = axislabels
  ) +
  geom_point(
    alpha = .25, size = 1, shape = 1
  ) +
  geom_abline(linewidth = .33, col = "dodgerblue") +
  theme(aspect.ratio = 1)

p_mood_biscatter_players <- p_mood_biscatter_sessions %+% 
  p_mood_difference_players$data

p_mood_biscatter <- (p_mood_biscatter_sessions / p_mood_biscatter_players)

p_mood_example / 
  (
    p_mood_prepost_sessions | 
      p_mood_prepost_players | 
      p_mood_difference | 
      p_mood_biscatter_players
  ) +
  plot_layout(heights = c(4.5, 5.5)) +
  plot_annotation(tag_levels = "A")
```

We then turned to our primary research question: To what extent does player mood in PWS change from pre-play to during play? First, we visualized the relevant data in @fig-data: Panel A shows mood responses from three example participants' first five sessions of play. @fig-data B then shows histograms of those data to facilitate visual comparison of the raw data. We then calculated sessions' and participants' mean moods at the beginning and during each session (@fig-data C, D, and G), and their differences (E and F). Visual inspection suggested greater mood during play compared to the pre-play measures at both the level of sessions and players.

@@

@@

```{r}
#| label: scaled-coefficients

posterior_pop <- as_draws_df(
  fit, 
  c("sd_pid__Intercept", "sd_pid__post1", "sigma", "b_post1")
) |> 
  mutate(
    total_residual = sd_pid__Intercept + sd_pid__post1 + sigma,
    b_post1_scaled = b_post1 / total_residual
  ) |> 
  tibble() |> 
  select(-starts_with("."))

posterior_pop_sum <- posterior_pop |> 
  summarise_draws(
    mean = ~number(mean(.), .001), 
    ~number2(quantile2(., probs = c(.025, .975)), .001)
  ) |> 
  mutate(
    x = str_glue("{mean} [{q2.5}, {q97.5}]")
  )
```

To further interpret the magnitude of the difference in moods between pre- and during play, we compared the contrast to the natural variations in mood as estimated by the model (standard deviation of the between-person intercept heterogeneity ($\tau_{\alpha_0}$ = `r posterior_pop_sum$x[1]`), between-person pre-during difference heterogeneity ($\tau_{\alpha_1}$ = `r posterior_pop_sum$x[2]`) and to the residual standard deviation ($\sigma$ = `r posterior_pop_sum$x[3]`). The standardized (with the total residual variation from the three above components) pre-play -- during play contrast was `r posterior_pop_sum$x[6]`, a not ignorable difference.
